\section{Introduction}
\label{sec:intro}

\paragraph{Who this is for.}
We assume \textbf{no prior coursework in cryptography}. You should be comfortable with binary, hexadecimal, and the idea that files are bytes; everything else---finite fields, S-boxes, block modes, and what ``AES'' actually does step by step---is built up in Section~\ref{sec:prelim} before we discuss the Python tracer in Section~\ref{sec:method}.

\paragraph{Why trace a toy cipher?}
In applications, encryption is often a single API call. That hides the internal \emph{round structure} that determines how quickly a one-bit change in plaintext or key \emph{avalanches} through the state. For learning (and for structured homework), we need \textbf{visibility}: intermediate 16-byte states after SubBytes, ShiftRows, MixColumns, and AddRoundKey, so we can plot Hamming distance along the pipeline and \textbf{see} diffusion.

\paragraph{What this document covers.}
We follow three threads that mirror a typical undergraduate lab brief and the \textbf{research-pipeline} style workflow used in \texttt{rd-ralph-template} (extract questions $\rightarrow$ design experiments $\rightarrow$ generate artifacts):
\begin{enumerate}
  \item \textbf{AES-128 internals} (Questions~1-style): a toy encrypt-only tracer, \texttt{toy\_aes128\_trace.py}, with the standard S-box, $\mathrm{GF}(2^8)$ arithmetic for MixColumns (\texttt{xtime}), optional ShiftRows ablation, and an alternate MixColumns matrix $M_{\mathrm{new}}$ for comparison.
  \item \textbf{Block cipher modes} (Questions~2-style): OpenSSL AES-128-ECB vs.\ AES-128-CBC on a BMP, illustrating why ECB leaks visual structure and CBC masks it at the byte level \citep{dworkin2007sp800-38a}.
  \item \textbf{TLS in the wild} (Questions~3-style): coarse negotiation snapshots with \texttt{openssl s\_client}, to be refined with Wireshark ClientHello inventories where the assignment requires them \citep{rfc5246}.
\end{enumerate}

\paragraph{Reproducibility.}
Every figure in Section~\ref{sec:experiments} can be regenerated by \texttt{python scripts/run\_coursework\_outputs.py}. The toy AES path is checked against PyCryptodome on a single block for standard options (\texttt{verify\_aes\_against\_pycryptodome}). This is \textbf{not} a production or side-channel-hardened implementation.
